---
title: "Coursera Capstone Project Report"
author: "Bright"
date: "`3/21/2025`"
output:
  pdf_document: default
  html_document: default
---

# Cyclistic Bike Share Usage Analysis

## Business Task Statement

To analyze how annual members and casual riders use Cyclistic bikes differently to identify key usage patterns and provide actionable recommendations to support strategic decision-making. 

## Data Sources

The data used for this analysis was provided by Motivate International Inc. under the data license agreement available at: <https://www.divvybikes.com/data-license-agreement>. This is publicly available data intended to explore customer behaviors and usage trends for Cyclistic bikes.

## Data Cleaning and Preparation Process 

To ensure the dataset was ready for accurate and meaningful analysis, the following steps were taken: 

#### Data Import and Consolidation 
  + Monthly datasets (12 Excel workbooks) were consolidated into a single dataset.
  + All files were verified for consistent column headers and data structure. 
  
#### Standardization and Cleaning 
  + Column headers were standardized. 
  + Data types were verified and corrected where necessary (e.g., date fields and numeric values).
  
#### New Calculated Columns 
  + Average Ride Length: Created to summarize the duration of each ride. 
  + Day of the Week: Extracted from ride start times using Excel's TEXT function for weekday analysis.
  
#### Pivot Charts and Data Exploration 
  + Monthly Trends: Analyzed average ride lengths by month and rider type. 
  + Weekday Patterns: Visualized ride frequency and duration by day of the week. 
  
#### Descriptive Statistics 
  + Computed minimum, maximum, and mode of ride lengths. 
  + Identified the mode day (most frequent ride day) for both rider types.
  
#### Data Consistency Checks 
  + Verified uniqueness of ride IDs and consistency of key fields like rideable type and timestamps.

#### Final Dataset Preparation 
  + Cleaned data with added insights was used for advanced analysis, 
forecasting, and pattern recognition.

## Summary of Analysis 

The cleaned data was imported into BigQuery for further analysis using SQL. Key findings include: 
  + Casual riders consistently exhibit longer average ride lengths than members. 
  + Casual ridership increases significantly in warmer months (March–July). 
  + Electrical bikes are the preferred rideable type for casual riders. 
  + Casual rides are more concentrated on weekends, while membership is evenly distributed throughout the week.
  
## Supporting Visualizations and Key Findings

The pivoted datasets were now imported into RStudio for further analysis and visualisation using R programming.

## Setting up my environment
Note: I set up my environment by installing and loading the 'tidyverse' and 'ggplot2' packages:

```{r echo=TRUE}
library(tidyverse)
library(ggplot2)
```

```{r load data, echo=TRUE}
average_rides <- readRDS("average_rides.rds")
```

```{r view dataframe}
view(average_rides)
```

## Average Ride Length by User Type

```{r viz Average ride length by user type, echo=TRUE}
ggplot(average_rides, aes(x = month, y = duration_minutes, fill = user_type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  labs(title = "Average Ride Duration Per Month",
           x = "Month",
           y = "Duration (minutes)",
           fill = "User Type") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1))
```

Based on the visualization;

  + Casual riders have a higher average ride length compared to members. 
  + Casual usage spikes seasonally, peaking during warmer months.

## Preferred Rideable Type

```{r load second data}
preferred_rides <- read.csv("Preferred Rideable Type.csv")
```

```{r}
preferred_rides <- rename(preferred_rides, 'total_rides' = Sum.of.total_rides)
```


```{r view second data}
view(preferred_rides)
```

```{r viz preferred_rides}
ggplot(preferred_rides, aes(x = ride_month, y = total_rides, fill = total_rides)) +
  geom_bar(stat = "identity") +
  labs(title = "Preferred Rideable Type",
           x = "Month",
           y = "Total Rides",
           fill = "Total Rides") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5),  # Center the title
    axis.text.x = element_text(angle = 0, hjust = 0.5)) # Center x-axis labels under bars
```

Based on the visualization;

  + Casual riders show a strong preference for electric bikes. 
  
## Number of Rides Per Day of the Week

```{r load third data}
rides_per_day <- read.csv("Number of Rides Per Day of the Week.csv")
```

```{r view third data}
view(rides_per_day)
```

```{r load scales to show number of rides}
library(scales)
```

```{r}
ggplot() +
  geom_bar(aes(x = rides_per_day$week - 0.2, y = rides_per_day$casual, fill = "Casual"),
           stat = "identity", width = 0.4) +
  geom_bar(aes(x = rides_per_day$week + 0.2, y = rides_per_day$member, fill = "Member"),
           stat = "identity", width = 0.4) +
  scale_fill_manual(values = c("Casual" = "skyblue", "Member" = "orange")) +
  scale_x_continuous(breaks = rides_per_day$week) +
  scale_y_continuous(labels = comma) +
  labs(title = "Daily Rides by User Type", x = "Day of the Week(Sunday - Saturday)", y = "Number of Rides", fill = "User Type") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

Based on the visualization;

  + Weekends see more casual ridership, while members show consistent weekday usage.

## Top Three Recommendations

#### Seasonal Promotions for Membership Conversion 
  + Offer targeted promotions (e.g., summer discounts, free trial months) to convert casual riders during peak seasons. 

#### Customized Membership Options 
  + Introduce flexible memberships (e.g., weekend-only, premium routes) tailored to casual rider preferences.
  
#### Incentives and Loyalty Programs 
  + Implement a points-based loyalty system for casual riders to earn rewards and discounts toward membership. 
  + Provide winter incentives like gear discounts to maintain engagement during low-use seasons.
  
  
This was a great learning journey for me.
All thanks to [Coursera](https://www.coursera.org/courseraplus?utm_medium=sem&utm_source=gg&utm_campaign=b2c_emea_x_coursera_ftcof_courseraplus_cx_dr_bau_gg_sem_bd-ex_s4_en_m_hyb_24-10_x&campaignid=21836581623&adgroupid=351685085710&device=c&keyword=coursera&matchtype=e&network=g&devicemodel=&creativeid=1449957450675&assetgroupid=&targetid=aud-2264666224030:kwd-36262515261&extensionid=&placement=&gad_source=1&gad_campaignid=21836581623&gbraid=0AAAAADdKX6bKWvqAjFQ2hedw3UI8Wy4_5&gclid=CjwKCAjwpMTCBhA-EiwA_-MsmX-Rl6xuKqfmyUmdeN81HA0o3uA_3JLYgru00cz0oRDS0PPW5ronXhoCOTQQAvD_BwE)
You can also learn new things. 
